Learning Discrete-valued Bayesian Networks from Mixed Data

### 2015-11-18 BN_discretizer_paper.tex

\begin{figure}[ht]
  \centering
  \begin{tabular}{cc}
    \input{graph1}
  \end{tabular}
  \caption{Factorization of $P(D \mid \Lambda)$}
  \label{fig:example_factorization}
\end{figure}

@@

\begin{algorithm}
\caption{Discretization of one continuous variable in a Bayesian network}
\label{alg:disc_one}
\begin{algorithmic}[5]
\Function{Discretize}{$D$, $G$}
\State
\State $N \leftarrow$ the number of instances
\State $N' \leftarrow$ the number of unique values in $D_X$
\State $H \leftarrow$ an $N \times N$ matrix such that $H[u,v] = h(u,v)$ per Algorithm~\ref{alg:h} in the appendix
\State $b \leftarrow$ the increasing sequence of $\Lambda$ integers as previously defined
\State $L \leftarrow$ the largest cardinality over all discrete variables in the Markov blanket of $X$
\State $S[u] \leftarrow$ the optimal objective value of subproblems with instances from $1$ to $u$
\State $M[u] \leftarrow$ the optimal discretization for subproblems with instances from $1$ to $u$
\State $W[u]  \leftarrow - \log\brock{1 - {\exp\paren{- L \cdot{ {{x_{b(u)+1} - x_{b(u)}}\over{x_N - x_1}}}}}}$ for $u = 1,2, \ldots,N'-1$ and $L(N') \leftarrow 0$
\State
\For {$v = 1$ to $N'$}
\If {$v = 1$}
\State $S[v] \leftarrow g \left(1,b[v] \right) + L[v]$
\State $M[v] \leftarrow \{ ({x_{b[v]} + x_{b[v]+1}}) / 2\}$
\Else
\State $s \leftarrow \infty$ and $boundary \leftarrow \infty$
\For {$u = 1$ to $v$}
\State $s' \leftarrow S[u] + g \left( b[u]+1,b[v] \right) +  {L \cdot {{x_{b[v]} - x_{b[u] + 1}}\over{x_N - x_1}}} + W[v]$
\If {$s' < s$}
\State $s \leftarrow s'$
\State $boundary \leftarrow ({x_{b[u]} + x_{b[u]+1}}) / 2$
\EndIf
\EndFor
\State $S[v] \leftarrow s$
\State $M[v] \leftarrow M[u] \cup \{ boundary\}$
\EndIf
\EndFor
 \State \Return $\Lambda$
\EndFunction
\end{algorithmic}
\end{algorithm}

@@

\begin{figure}[ht]
     %\begin{tabular}{cc}
      \input{graph_housing_exp2_distr_3_5}
     %\end{tabular}
  \caption{The discretization result on variable $C$ and $E$ of the network in Figure~\ref{fig:housing_graph_2}. The high number of discretization edges is due to the repeated values on the dataset in variable $C$ and $E$.}
  \label{fig:housing_exp2_distr_3_5}
\end{figure}

\begin{figure}[ht]
    % \begin{tabular}{cc}
      \input{graph_housing_exp2_distr_8_5}
    % \end{tabular}
  \caption{The discretization result on variable $H$ and $E$ of the network in Figure~\ref{fig:housing_graph_2}. Although the number of discretization edges on $H$ is not as large as $E$'s, they still match the distribution of the raw data very well.}
  \label{fig:housing_exp2_distr_8_5}
\end{figure}

\subsection{Experiment Discussions}
\label{subsec:discuss_exp}

\begin{figure}[ht]
    % \begin{tabular}{cc}
      \input{graph_exp_discuss}
    % \end{tabular}
  \caption{A simple Bayesian network that is used to demonstrate the sensitivity of the three methods}
  \label{fig:exp_discuss}
\end{figure}

To quantatively discuss the sensitivity of the three methods, consider the Bayesian network in Figure \ref{fig:exp_discuss}, where $X$ is continuous-valued and $P_1$ and $P_2$ is discrete-valued. Let $P = \{  P_1, P_2 \}$. Assume $D_X = \{1.0,2.0,3.0, \ldots N \}$. Therefore, the corresponding objective functions to discretize $X$ in the three methods are:
\begin{equation}
\begin{aligned}
\text{Obj}_{\text{MDL}} & = Z_1 \cdot k + \log(N) \cdot \sum^k_{i=1} \sum^{J_P}_{j=1} I(a^*_i, (p_1,p_2)_j) \\
\text{Obj}_{\text{Bayesian}} &= Z_2 \cdot k + \log {{n_i + J_P - 1}\choose{J_P -1}} + \sum^k_{i=1} \log \left(   {{n_i !}\over{n^{(P)}_{i,1}! n^{(P)}_{i,2}! \cdots n^{(P)}_{i,J_P}!    }} \right) \\
\text{Obj}_{\text{Heuristic}} &= Z_2 \cdot k + \log {{n_i + J_{P_1} - 1}\choose{J_{P_1} -1}} + \log {{n_i + J_{P_2} - 1}\choose{J_{P_2} -1}} +\\
& \sum^k_{i=1}  \left[ \log \left(   {{n_i !}\over{n^{(P_1)}_{i,1}! n^{(P_1)}_{i,2}! \cdots n^{(P_1)}_{i,J_{P_1}}    }} \right)  + \log \left(   {{n_i !}\over{n^{(P_2)}_{i,1}! n^{(P_2)}_{i,2} \cdots n^{(P_2)}_{i,J_{P_2}} }} \right) \right]
\end{aligned}
\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{acknowledgements}
%If you'd like to thank anyone, place your comments here
%and remove the percent signs.
%\end{acknowledgements}

>>>>

\begin{figure}[ht]
  \centering
  \input{graph1}
  \caption{Factorization of $P(D \mid \Lambda)$}
  \label{fig:example_factorization}
\end{figure}

@@

\begin{algorithm}
\caption{Discretization of one continuous variable in a Bayesian network}
\label{alg:disc_one}
\begin{algorithmic}[5]
\Function{Discretize}{$D$, $G$}
\State $N \leftarrow$ the number of instances
\State $N' \leftarrow$ the number of unique values in $D_X$
\State $H \leftarrow$ an $N \times N$ matrix such that $H[u,v] = h(u,v)$ per Algorithm~\ref{alg:h} in the appendix
\State $b \leftarrow$ the increasing sequence of $\Lambda$ integers as previously defined
\State $L \leftarrow$ the largest cardinality over all discrete variables in the Markov blanket of $X$
\State $S[u] \leftarrow$ the optimal objective value of subproblems with instances from $1$ to $u$
\State $M[u] \leftarrow$ the optimal discretization for subproblems with instances from $1$ to $u$
\State $W[u]  \leftarrow - \log\brock{1 - {\exp\paren{- L \cdot{ {{x_{b(u)+1} - x_{b(u)}}\over{x_N - x_1}}}}}}$ for $u = 1,2, \ldots,N'-1$ and $L(N') \leftarrow 0$
\For {$v \leftarrow 1$ to $N'$}
\If {$v = 1$}
\State $S[v] \leftarrow g \left(1,b[v] \right) + L[v]$
\State $M[v] \leftarrow \{ ({x_{b[v]} + x_{b[v]+1}}) / 2\}$
\Else
\State $s \leftarrow \infty$ and $boundary \leftarrow \infty$
\For {$u \leftarrow 1$ to $v$}
\State $s' \leftarrow S[u] + g \left( b[u]+1,b[v] \right) +  {L \cdot {{x_{b[v]} - x_{b[u] + 1}}\over{x_N - x_1}}} + W[v]$
\If {$s' < s$}
\State $s \leftarrow s'$
\State $boundary \leftarrow ({x_{b[u]} + x_{b[u]+1}}) / 2$
\EndIf
\EndFor
\State $S[v] \leftarrow s$
\State $M[v] \leftarrow M[u] \cup \{ boundary\}$
\EndIf
\EndFor
 \State \Return $\Lambda$
\EndFunction
\end{algorithmic}
\end{algorithm}

@@

\begin{figure}[ht]
\centering
     %\begin{tabular}{cc}
      \input{graph_housing_exp2_distr_3_5}
     %\end{tabular}
  \caption{The discretization result on variable $C$ and $E$ of the network in Figure~\ref{fig:housing_graph_2}. The high number of discretization edges is due to the repeated values on the dataset in variable $C$ and $E$.}
  \label{fig:housing_exp2_distr_3_5}
\end{figure}

\begin{figure}[ht]
\centering
    % \begin{tabular}{cc}
      \input{graph_housing_exp2_distr_8_5}
    % \end{tabular}
  \caption{The discretization result on variable $H$ and $E$ of the network in Figure~\ref{fig:housing_graph_2}. Although the number of discretization edges on $H$ is not as large as $E$'s, they still match the distribution of the raw data very well.}
  \label{fig:housing_exp2_distr_8_5}
\end{figure}

\subsection{Discussions}
\label{subsec:discuss_exp}

\begin{figure}[ht]
\centering
    % \begin{tabular}{cc}
      \input{graph_exp_discuss}
    % \end{tabular}
  \caption{A simple Bayesian network that is used to demonstrate the sensitivity of the three methods}
  \label{fig:exp_discuss}
\end{figure}

To quantatively discuss the sensitivity of the three methods, consider the Bayesian network in Figure \ref{fig:exp_discuss}, where $X$ is continuous-valued and $P_1$ and $P_2$ is discrete-valued. Let $\Pi_X = \{  P_1, P_2 \}$. Assume $D_X = \{1.0,2.0,3.0, \ldots N \}$. Therefore, the corresponding objective functions to discretize $X$ in the three methods are:
{\begin{small}
\begin{equation}
\begin{aligned}
\text{Obj}_{\text{MDL}} & = \overbrace{Z_1 \cdot k + \log(k) + \log {{N +k - 1}\choose{k -1}} }^{\text{Penalty Term}}+   \overbrace{\log(N) \cdot I(X^*, \Pi_X)}^{\text{Disc. Edge Positions Term}} \\
\text{Obj}_{\text{Bayesian}} &= \overbrace{Z_2 \cdot k + \log {{n_i + J_P - 1}\choose{J_P -1}}}^{\text{Penalty Term}} + \overbrace{\sum^k_{i=1} \log \left(   {{n_i !}\over{n^{(P)}_{i,1}! n^{(P)}_{i,2}! \cdots n^{(P)}_{i,J_P}!    }} \right) }^{\text{Disc. Edge Positions Term}}\\
\text{Obj}_{\text{Heuristic}} &= \overbrace{Z_2 \cdot k + \log {{n_i + J_{P_1} - 1}\choose{J_{P_1} -1}} + \log {{n_i + J_{P_2} - 1}\choose{J_{P_2} -1}} }^{\text{Penalty Term}}+\\
& \overbrace{\sum^k_{i=1}  \left[ \log \left(   {{n_i !}\over{n^{(P_1)}_{i,1}! n^{(P_1)}_{i,2}! \cdots n^{(P_1)}_{i,J_{P_1}}    }} \right)  + \log \left(   {{n_i !}\over{n^{(P_2)}_{i,1}! n^{(P_2)}_{i,2} \cdots n^{(P_2)}_{i,J_{P_2}} }} \right) \right]}^{\text{Disc. Edge Positions Term}}
\end{aligned}
\end{equation}
\end{small}}
, where $Z_1$ and $Z_2$ are two constants, $k$ is the number of intervals after discretization, $X^*$ is the discretized version of $X$, and $I(X^*, \Pi_X)$ is the mutual information between $X^*$ and its parent variables. Note that the third term in $\text{Obj}_{\text{MDL}}$ is approximated by $H(p)$ as in Equation \ref{eqn:MDL} in the original paper \citep{Friedman_1996}.

The sensitivities toward the value distribution in $D_{\Pi_X}$ are determined by the `` Disc. Edge Positions Term" in the three objective functions. Since the penaltis of the numbers of intervals after discretization in the three objective functions take the similar form, if the `` Disc. Edge Positions Term" varies more severely due to different positions of discretization edges, then the corresponding method is more sensitive. The `` Disc. Edge Positions Term" in $\text{Obj}_{\text{MDL}}$, which is made out of the mutual information, varies less severely than the terms in $\text{Obj}_{\text{Bayesian}}$ and $\text{Obj}_{\text{Heuristic}}$, since it senses the distribution in $D_{\Pi_X}$ by empirical probability distributions, instead of the factorial terms such in $\text{Obj}_{\text{Bayesian}}$ and $\text{Obj}_{\text{Heuristic}}$. Furthermore, the `` Disc. Edge Positions Term" in $\text{Obj}_{\text{Heuristic}}$ varies more aggressively than the term in $\text{Obj}_{\text{Bayesian}}$. Therefore, the MDL method is less sensitive and the heuristic is over sensitive toward the value distributions in other variables. The effect of sensitivity can be shown as the numbers of intervals after discretization in the previous experiment.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
In this work, a new method for discretizating continuous varibles in Bayesian networks has been proposed. It is compatible with the K2 structure learning, i.e., it can discretize continuous variables along with the edge-adding process of the K2. Therefore, a discrete-valued Bayesian network can be learned from mix data.

Experiments on three real-world dataset show that the new method returns to a better discretization result than previous research in terms of likelihood. The discretization edges and the resulting marginal probabilities of discretized regions returned by the new method can match the row data well.

This work can extend the usage of Bayesian networks in science and engineering, especially for the fields that require discrete-valued variables but the learning data is mixed and numerical. In the future, we plan to generalize the work to the datast that has high dimensional variables, for example, 3 dimensional positions of drones, which makes the sorting in Equation \ref{eq:p_M} fails.

%\begin{acknowledgements}
%If you'd like to thank anyone, place your comments here
%and remove the percent signs.
%\end{acknowledgements}
